MISUSES AND INACCURACIES OF PAGEL'S LAMBDA ML ESTIMATION METHODS

### 2020-04-21 Manuscript/Manuscript-1-DCA.tex
vsn 1 edits

For many macroevolutionary analyses, it is often of interest to quantify
the degree to which phylogenetic signal is displayed in continuous
traits. Several analytical tools have been developed for this purpose
(e.g., Gittleman and Kot 1990; Abouheif 1999; Pagel 1999; Blomberg et
al. 2003; Klingenberg and Gidaszewski 2010; Adams 2014a), which differ
primarily in how they characterize the phylogenetic dependency of trait
variation among taxa. One commonly used statistical measure,
\emph{Kappa} (\(K\)), expresses the strength of phylogenetic signal as
the ratio of observed trait variation to the trait variation conditioned
on the phylogeny; scaled by what is expected under Brownian motion given
the phylogeny's size and shape (Blomberg et al. 2003; also Adams 2014a).
Another approach, Pagel's \(\lambda\) (Pagel 1999), uses maximum
likelihood to fit the data to the phylogeny under some model of
evolutionary change (typically Brownian motion). The inclusion of a
scaling parameter, \(\lambda\), transforms the lengths of the internal
branches of the phylogeny to improve the fit, and this parameter
describes the degree of phylogenetic signal in the dataset (Pagel 1999;
Freckleton et al. 2002). Pagel's \(\lambda\) also has the advantage that
it may be included when estimating the association of traits in a
phylogenetic context, meaning that one may account for the degree of
phylogenetic signal while conducting phylogenetic regression or ANOVA
(see Freckleton et al. 2002). \hfill\break

Several studies have investigated the statistical properties of methods
for estimating phylogenetic signal under various conditions (e.g.,
Munkemuller et al. 2012; Pavoine and Ricotta 2012; Diniz-Filho et al.
2012; Molina-Venegas and Rodriguez 2017; see also Revell et al. 2008;
Revell 2010). These have largely focused on the ability of methods to
detect the presence of phylogenetic signal (i.e., type I and type II
error rates) under complex models of evolutionary change, across a range
of phylogeny sizes, and with varying degrees of phylogenetic uncertainty
or unresolved topologies. In terms of parameter estimation, Revell
(2010) found that regression parameters were accurately estimated when
\(\lambda\) was included during phylogenetic regression, and Munkmuller
et al.~(2012) demonstrated that estimates of phylogenetic signal
obtained using various measures generally increased when input levels of
phylogenetic signal were stronger. However, the precision of those
estimates could not be determined, because the input levels of
phylogenetic signal were simulated via a scaling factor (\(w\)) that was
not directly comparable to the measures of phylogenetic signal being
compared (see Munkemuller et al. 2012). One study (Boettiger et al.
2012) found that estimates of Pagel's \(\lambda\) displayed less
variation when data were simulated on a large phylogeny (\(N=281\)) as
compared to a small one (\(N=13\)), concluding that insufficient data
(i.e., the number of species) was the underlying cause of the lack of
precision in parameter estimation. However, this conclusion assumes that
the precision in estimating \(\lambda\) remains constant across its
range (\(\lambda = 0 \to 1\)); an assumption that to date, has not been
verified. Thus, despite widespread use of Pagel's (1999) \(\lambda\) in
macroevolutionary studies, at present, we still lack a general
understanding of the precision with which \(\lambda\) can estimate
levels of phylogenetic signal in phenotypic datasets. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. First
we use computer simulations across differing numbers of species,
differently shaped phylogenies, and differing input levels of
phylogenetic signal, to explore the degree to which \(\lambda\)
correctly identifies known levels of phylogenetic signal, and under what
circumstances. For comparison, we also evaluate estimates of \(Kappa\)
and its empirically-derived effect size (\(Z_K\)) for this purpose.
Additionally, we use simulations to determine how the inclusion of
\(\lambda\) in phylogenetic regression and ANOVA (i.e., PGLS) affects
parameter estimation, and whether levels of phylogenetic signal
estimated in a PGLS framework are accurate. We then survey the recent
macroevolutionary literature for published papers containing estimates
of \(\lambda\) from empirical datasets, and compare these empirical
estimates to patterns gleaned from our computer simulations. In general
we find that while PGLS parameters (e.g., \(\beta\)) are accurately
estimated with the inclusion of phylogenetic signal, estimates of
\(\lambda\) are not. Further, we find that estimates of \(\lambda\) vary
widely for a given input value of phylogenetic signal, and that this
variation is not constant across the range of input signal, with
decreased precision when phylogenetic signal is of intermediate
strength. By contrast, variation across effect sizes (\(Z_K\)) obtained
from \(Kappa\) is far more consistent across the range of input values,
and the precision of estimating \(Z_K\) is preferable to that observed
using \(\lambda\). We conclude that estimates of phylogenetic signal
using Pagel's \(\lambda\) are often innacurate, and thus interpreting
strength of phylogenetic signal in phenotypic datasets based on this
measure should be treated with caution. By contrast, effect sizes
obtained from \(Kappa\) hold promise for characterizing phylogenetic
signal, and comparing the strength of phylogenetic signal across
datasets.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in
each dataset using Pagel's \(\lambda\). Next, for the 1050 datasets at
each level of species richness (\(n\)), we assessed the relationship
between \(\lambda_{in}\) and \(\lambda_{est}\) using regression.
Additionally, the precision of \(\lambda\) was approximated by observing
the variation of \(\lambda_{est}\) obtained at each input level of
phylogenetic signal (\(\lambda_{in}\)). Finally, for comparison we
estimated the degree of phylogenetic signal in each dataset using
\(Kappa\) (Blomberg et al. 2003; Adams 2014a), and from this calculated
its effect size (\(Z_K\): sensu Adams and Collyer 2016, 2019a) for each
dataset. Here, the effect size was estimated as:
\(Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}\), where \(K_{obs}\) was the
observed \(Kappa\), and \(\mu_K\) and \(\sigma_K\) were the mean and
standard deviation of the empirical sampling distribution of values
obtained from the permutation distribution (for discussion see: Adams
and Collyer 2019b). Because \(Z_K\) differs in scale from \(\lambda\),
we used a linear normalization to standardize \(Z_K\) to a uniform
distribution (\(0\rightarrow1\)), and variation in the set of normalized
\(Z_K\) values at each input level of phylogenetic signal
(\(\lambda_{in}\)) were then estimated for comparison. \hfill\break

@@

To determine how Pagel's \(\lambda\) is commonly utilized in empirical
studies, we conducted a literature survey. From Google.scholar we
obtained a list of all papers published in 2019 that used \(\lambda\);
resulting in 341 studies. For each study, we extracted all
\(\lambda_{est}\), the size of the phylogeny (\(n\)), and noted whether
authors reported confidence intervals or performed significance tests
assessing difference of \(\lambda_{est}\) from either zero or one. We
also noted whether biological interpretations based on \(\lambda_{est}\)
were made, and for studies that reported more than one
\(\lambda_{est}\), we also noted whether these were compared in some
manner, and whether such comparisons were accompanied with statistical
tests between \(\lambda_{est}\). \hfill\break

@@

Our primary series of simulations revealed several patterns. First, the
relationship between \(\lambda_{est}\) and input levels of phylogenetic
signal was \(\sim\beta=1.0\) at each level of species richness
(\(n=32, 64, 128, 256, 512, 1024\)). This implied that the mean
\(\lambda_{est}\) across datasets for a given set of simulation
conditions reflected input levels of phylogenetic signal. However, as
shown in Figure 1, the precision of those estimates varied widely, and
in several interesting ways. Predictably, precision was worst at low
levels of species richness, where in many cases the set of
\(\lambda_{est}\) spanned nearly the entire range of possible values
(e.g., \(n=32\); \(\lambda_{est}=0.5\): range = \(0.0\to0.985\)).
Second, as species richness increased, variation across estimates of
\(\lambda\) decreased (Figure 1). This confirmed the pattern identified
by Boettiger et al.~(2012) on a small (\(n=13\)) and large (\(n=281\))
phylogeny, and demonstrated that this trend was general across a wider
range of species richness. Importantly however, our broader set of
simulations revealed that the precision of \(\lambda_{est}\) was not
uniform across all levels of phylogenetic signal, and that there was a
particular pattern to the imprecision of estimating this phylogenetic
parameter. Specifically, we found that variation in \(\lambda_{est}\)
was highest at intermediate levels of phylogenetic signal, and decreased
at both lower and higher levels of input signal (Figure 1). This implied
that the precision in estimating \(\lambda\) was worst at intermediate
levels of phylogenetic signal; a pattern that remained consistent across
all levels of species richness evaluated here (Figure 1). Notably, even
at very large levels of species richness, the range of \(\lambda_{est}\)
still encompassed a substantial portion of possible values (e.g.,
\(n=1024\); \(\lambda_{est}=0.5\): range = \(0.31\to 0.68\)). From this
we conclude that \(\lambda_{est}\) is not an accurate estimator of the
strength of phylogenetic signal in phenotypic datasets, and that
biological interpretations of the level of phylogenetic signal based on
\(\lambda\) should be treated with caution. \hfill\break

@@

By contrast, (FIg 2) Kappa. Much smaller range and variation as compared
with lambda, but with larger l-in, variation in kappa does increase.
However, this is not the case with Z-K. Z-K increases predictably
(linearly, progressively\ldots) with L-in, and importantly variation in
Z-K remains nearly constant across the range of input values. this means
that Z-K is equally precise in estimating the strength of phylogenetic
signal, and may be more adept at facilitating useful statements on teh
relative strength of phylogenetic signal within and across datasets.
\hfill\break

\textbf{NOTE: Fig 2 is 4 panels for N=128: Lambda, Kappa, Z-K,
Z-k-normalized.}

@@

@@

{[}insert Figure 3 here{]} \hfill\break

>>>>

For many macroevolutionary analyses, it is often of interest to quantify
the degree to which phylogenetic signal is displayed in continuous
traits. Several analytical tools have been developed for this purpose
(e.g., Gittleman and Kot 1990; Abouheif 1999; Pagel 1999; Blomberg et
al. 2003; Klingenberg and Gidaszewski 2010; Adams 2014a), which differ
primarily in how they characterize the phylogenetic dependency of trait
variation among taxa. One commonly used statistical measure,
\emph{Kappa} (\(K\)), expresses the strength of phylogenetic signal as
the ratio of observed trait variation to the trait variation conditioned
on the phylogeny; scaled by what is expected under Brownian motion given
the phylogeny's size and shape (Blomberg et al. 2003; also Adams 2014a).
Another approach, Pagel's \(\lambda\) (Pagel 1999), uses maximum
likelihood to fit the data to the phylogeny under some model of
evolutionary change (typically Brownian motion). The inclusion of a
scaling parameter, \(\lambda\), transforms the lengths of the internal
branches of the phylogeny to improve the fit, and this parameter
describes the degree of phylogenetic signal in the dataset (Pagel 1999;
Freckleton et al. 2002). Pagel's \(\lambda\) also has the appeal that it
may be included when estimating the association of traits in a
phylogenetic context, meaning that one may account for the degree of
phylogenetic signal while conducting phylogenetic regression or ANOVA
(see Freckleton et al. 2002). \hfill\break

Several studies have investigated the statistical properties of methods
for estimating phylogenetic signal under various conditions (e.g.,
Munkemuller et al. 2012; Pavoine and Ricotta 2012; Diniz-Filho et al.
2012; Molina-Venegas and Rodriguez 2017; see also Revell et al. 2008;
Revell 2010). These have largely focused on the ability of methods to
detect the presence of phylogenetic signal (i.e., type I and type II
error rates) under complex models of evolutionary change, across a range
of phylogeny sizes, and with varying degrees of phylogenetic uncertainty
or unresolved topologies. In terms of parameter estimation, Revell
(2010) found that regression parameters were accurately estimated when
\(\lambda\) was included during phylogenetic regression, and Munkmuller
et al.~(2012) demonstrated that estimates of phylogenetic signal
obtained using various measures generally increased when input levels of
phylogenetic signal were stronger. However, the precision of those
estimates could not be determined, because the input levels of
phylogenetic signal were simulated via a scaling factor (\(w\)) that was
not directly comparable to the measures of phylogenetic signal being
compared (see Munkemuller et al. 2012). One study (Boettiger et al.
2012) found that estimates of Pagel's \(\lambda\) displayed less
variation when data were simulated on a large phylogeny (\(N=281\)) as
compared to a small one (\(N=13\)), concluding that insufficient data
(i.e., the number of species) was the underlying cause of the increased
variation across parameter estimates. However, this conclusion assumes
that the precision in estimating \(\lambda\) remains constant across its
range (\(\lambda = 0 \to 1\)); an assumption that to date, has not been
verified. Thus, despite widespread use of Pagel's (1999) \(\lambda\) in
macroevolutionary studies, at present, we still lack a general
understanding of the precision with which \(\lambda\) can estimate
levels of phylogenetic signal in phenotypic datasets. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We also
evaluate the precision of a standardized effect size (\(Z_K\)) based on
\(Kappa\) for comparison. Additionally, we use simulations to determine
how the inclusion of \(\lambda\) in phylogenetic regression and ANOVA
(i.e., PGLS) affects parameter estimation, and whether levels of
phylogenetic signal estimated in a PGLS framework are accurate. We then
survey the recent macroevolutionary literature for published papers
containing estimates of \(\lambda\) from empirical datasets, and compare
these empirical estimates to patterns gleaned from our computer
simulations. In general we find that while PGLS parameters (e.g.,
\(\beta\)) are accurately estimated with the inclusion of phylogenetic
signal, estimates of \(\lambda\) are not. We also find that estimates of
\(\lambda\) vary widely for a given input value of phylogenetic signal,
and that this variation is not constant across the range of input
signal, with decreased precision when phylogenetic signal is of
intermediate strength. In addition, the same \(\lambda_{est}\) may be
obtained from datasets containing vastly different input levels of
phylogenetic signal. By contrast, variation across effect sizes
(\(Z_K\)) obtained from \(Kappa\) is far more consistent across the
range of input values, and the precision of \(Z_K\) is generally
preferable to that of \(\lambda_{est}\). Further, because \(Z_K\) is a
standardized effect size, the strength of phylogenetic signal is more
readily interpretable, and may be compared across datasets. We conclude
that estimates of phylogenetic signal using Pagel's \(\lambda\) are
often innacurate, and thus interpreting strength of phylogenetic signal
in phenotypic datasets based on this measure is compromised. By
contrast, effect sizes obtained from \(Kappa\) hold promise for
characterizing phylogenetic signal, and for comparing the strength of
phylogenetic signal across datasets.

@@

Using the simulated phylogenies we estimated the phylogenetic signal in
each dataset using Pagel's \(\lambda\). Next, for the 1050 datasets at
each level of species richness (\(n\)), we assessed the relationship
between \(\lambda_{in}\) and \(\lambda_{est}\) using regression.
Additionally, the precision of \(\lambda\) was approximated by observing
the variation of \(\lambda_{est}\) obtained at each input level of
phylogenetic signal (\(\lambda_{in}\)). Finally, for comparison we
characterized the strength of phylogenetic signal in each dataset using
a standardized effect size (\(Z_K\): sensu Adams and Collyer 2016,
2019a) based on \(Kappa\). As suggested by Adams and Collyer (2019b),
the effect size for phylogenetic signal may be estimated as:
\(Z_K=\frac{K_{obs}-\mu_K}{\sigma_K}\), where \(K_{obs}\) was the
observed \(Kappa\), and \(\mu_K\) and \(\sigma_K\) were the mean and
standard deviation of the empirical sampling distribution of values
obtained from the permutation distribution. \(Z_K\) describes the
strength of phylogenetic signal as a standard deviate from its sampling
distribution, and thus directly measures the strength of signal in a
manner that is comparable across datasets. Variation in the set of
\(Z_K\) at each input level of phylogenetic signal was then calculated
as an estimate of precision in \(Z_K\). However, because \(Z_K\) differs
in scale from \(\lambda\), we used a linear normalization to standardize
\(Z_K\) to a uniform distribution (\(0\rightarrow1\)), and estimated the
precision of \(Z_K\) from the normalized values. \hfill\break

@@

To determine how Pagel's \(\lambda\) is commonly utilized in empirical
studies, we conducted a literature survey. From Google.scholar we
obtained a list of all papers published in 2019 that used \(\lambda\);
resulting in 341 studies. For each study, we extracted all
\(\lambda_{est}\), the size of the phylogeny (\(n\)), and noted whether
authors reported confidence intervals or performed significance tests
assessing difference of \(\lambda_{est}\) from either zero or one. We
also noted whether biological interpretations based on \(\lambda_{est}\)
were made, and for studies that reported more than one
\(\lambda_{est}\), we also noted whether these were compared in some
manner, and whether such comparisons were accompanied with statistical
tests between \(\lambda_{est}\).

@@

Our first set of simulations revealed several patterns. First, the
relationship between \(\lambda_{est}\) and \(\lambda_{in}\) across the
range input levels of phylogenetic signal was \(\sim\beta=1.0\);
indicating that the average \(\lambda_{est}\) across datasets for a
given set of simulation condition correctly reflected input levels of
phylogenetic signal. However, as shown in Figure 1, the precision of
those estimates varied widely, and in several interesting ways.
Predictably, precision was worse at low levels of species richness,
where in many cases the set of \(\lambda_{est}\) spanned nearly the
entire range of possible values (e.g., \(n=32\); \(\lambda_{in}=0.5\):
range of \(\lambda_{est}= 0.0\to 0.985\)). Second, as species richness
increased, variation across estimates of \(\lambda\) decreased (Figure
1). This confirmed the pattern identified by Boettiger et al.~(2012) on
a small (\(n=13\)) versus a large (\(n=281\)) phylogeny, and
demonstrated that the trend of increasing precision with higher species
richness was general. Importantly however, our broader set of
simulations revealed that the precision of \(\lambda_{est}\) was not
uniform across all levels of phylogenetic signal. Specifically, we found
that variation in \(\lambda_{est}\) was highest at intermediate levels
of phylogenetic signal, and decreased at both lower and higher levels of
input signal (Figure 1). This implied that the precision in estimating
\(\lambda\) was worse at intermediate values, and improved as the levels
of phylogenetic signal were closer to \(\lambda_{in}=0\) (no
phylogenetic signal) or \(\lambda_{in}=1\) (Brownian motion). Notably,
even at large levels of species richness, the range of \(\lambda_{est}\)
still encompassed a substantial portion of possible values (e.g.,
\(n=512\); \(\lambda_{in}=0.5\): range of
\(\lambda_{est} = 0.32\to 0.68\)). Likewise, the same \(\lambda_{est}\)
could be obtained from datasets containing vastly different input levels
of phylogenetic signal (e.g., \(n=512\); \(\lambda_{est} = 0.5\); range
of \(\lambda_{in} = 0.25 \to 0.65\)). Taken together, these findings
reveal that \(\lambda_{est}\) does not precisely characterize observed
levels of phylogenetic signal in phenotypic datasets, and that
biological interpretations of the strength of phylogenetic signal based
on \(\lambda\) may be highly inaccurate. \hfill\break

@@

By contrast, when characterizing phylogenetic signal with the
standardized effect size (\(Z_K\)) of \(Kappa\), we found that the
precision of \(Z_K\) was more stable, as variation across datasets was
far more consistent across the range of input values. For example, when
\(n=128\) the precision of \(\lambda_{est}\) (Figure 2A) varied
considerably more across input levels of phylogenetic signal than did
the precision of \(Z_K\) for the same datasets (Figure 2B). Further, for
a given \(n\), the variance in precision estimates of \(Z_K\) was
considerably smaller than the variance in precision estimates of
\(\lambda_{est}\) (Fig. 2C,D); implying that estimates of phylogenetic
signal were more reliable and robust when using \(Z_K\) (for additional
results see Supporting Information). Finally, it should also be
recognized that because \(Z_K\) is a standardized effect size, the
strength of phylogenetic signal is more readily interpretable when using
this measure, as \(Z_K\) expresses the strength of phylogenetic signal
in standard deviation units relative to the mean (see Adams and Collyer
2019b). This further implies that comparisons of the strength of
phylogenetic signal among phenotypic traits are possible, and may be
accomplished statistically via a two-sample test that formally compares
\(Z_K\) across datasets (for comparisons of multivariate effect sizes
see: Adams and Collyer 2016, 2019a). \hfill\break

@@

When \(\lambda\) was incorporated in PGLS regression and ANOVA (i.e.,
\(Y\sim{X}\)), we found much the same pattern as in the previous
simulations. Namely, the precision of \(\lambda_{est}\) varied inversely
with species richness; where greater precision was obtained at higher
levels of species richness (Figure 3). Likewise, the precision in
estimating \(\lambda\) was worse at intermediate values (Figure 3), and
improved as the levels of phylogenetic signal were closer to
\(\lambda_{in}=0\) (no phylogenetic signal) or \(\lambda_{in}=1\)
(Brownian motion). Generally, the spread of estimates was slightly
broader when \(\lambda\) was co-estimated with regression parameters in
PGLS regression (Figure 3), as compared to when \(\lambda\) was
estimated in cases only the dependent variable (Y) was considered
(Figure 1). Finally, found that regression parameters (\(\beta\)) were
accurately estimated when \(\lambda\) was included during phylogenetic
regression; a result which confirmed earlier findings of Revell (2010)
(see Supporting Information). \hfill\break

{[}insert Figure 3 here{]}

@@

{[}insert Figure 4 here{]} \hfill\break
